---
title: "Baseline"
format: html
---

## Goal
Establish a baseline performance profile using logistic regression and observe how common evaluation metrics behave before introducing model complexity.

```{r}
source("../R/config.R")
library(tidyverse)
library(rsample)
library(yardstick)
library(broom)
```

```{r}
train_df <- readRDS("../data/modeling_train.rds")
test_df  <- readRDS("../data/modeling_test.rds")

glimpse(train_df)
```

```{r}
train_df <- train_df %>% mutate(default = factor(default))
test_df  <- test_df  %>% mutate(default = factor(default))

train_df <- train_df %>%
  mutate(default = relevel(default, ref = "0"))

baseline_model <- glm(
  default ~ .,
  data = train_df,
  family = binomial()
)

summary(baseline_model)
```

```{r}
test_probs <- predict(baseline_model, newdata = test_df, type = "response")

baseline_preds <- test_df %>%
  mutate(
    .pred = as.numeric(test_probs),
    .class = if_else(.pred >= 0.5, "1", "0") %>%
    factor(levels = levels(default))
  )
```

```{r}
ev <- "second"

baseline_results <- bind_rows(
  accuracy(baseline_preds, truth = default, estimate = .class),
  roc_auc(baseline_preds, truth = default, .pred, event_level = ev),
  pr_auc(baseline_preds, truth = default, .pred, event_level = ev),
  mn_log_loss(baseline_preds, truth = default, .pred, event_level = ev),
  brier_class(baseline_preds, truth = default, .pred, event_level = ev)
) %>%
  mutate(model = "Baseline Logistic")

baseline_results
```

## Baseline Results

This logistic regression model establishes a reference performance profile. At this stage, most metrics are expected to move together since the model is simple and relatively well-calibrated.

These values will serve as the comparison point for more complex models in the next notebook.

