---
title: "Baseline"
format: html
editor: visual
---

## Goal
Establish a baseline performance profile using logistic regression and observe how common evaluation metrics behave before introducing model complexity.

```{r}
source("../R/config.R")
library(tidyverse)
library(rsample)
library(yardstick)
library(broom)
```

```{r}
train_df <- readRDS("../data/modeling_train.rds")
test_df  <- readRDS("../data/modeling_test.rds")

glimpse(train_df)
```

```{r}
train_df <- train_df %>% mutate(default = factor(default))
test_df  <- test_df  %>% mutate(default = factor(default))

baseline_model <- glm(
  default ~ .,
  data = train_df,
  family = binomial()
)

summary(baseline_model)
```

```{r}
test_probs <- predict(baseline_model, test_df, type = "response")

baseline_preds <- test_df %>%
  mutate(
    .pred = test_probs,
    .class = if_else(.pred >= 0.5, "1", "0") %>% factor(levels = levels(default))
  )
```

```{r}
metrics_class <- metric_set(
  accuracy
)

class_results <- metrics_class(
  baseline_preds,
  truth = default,
  estimate = .class
)

metrics_prob <- metric_set(
  roc_auc,
  pr_auc,
  mn_log_loss,
  brier_class
)

prob_results <- metrics_prob(
  baseline_preds,
  default,
  .pred
)

baseline_results <- bind_rows(class_results, prob_results)
baseline_results

```

## Baseline Results

This logistic regression model establishes a reference performance profile. At this stage, most metrics are expected to move together since the model is simple and relatively well-calibrated.

These values will serve as the comparison point for more complex models in the next notebook.

