---
title: "Complex Models"
format: html
---

## Goal

Evaluate how common metrics behave when model complexity increases and models are tuned specifically to optimize ROC-AUC.

```{r}
source("../R/config.R")

library(tidyverse)
library(rsample)
library(yardstick)
library(broom)
library(ranger)
library(xgboost)
library(glmnet)
library(tidyr)
library(ggplot2)
```

```{r}
train_df <- readRDS("../data/modeling_train.rds")
test_df  <- readRDS("../data/modeling_test.rds")

train_df <- train_df %>% mutate(default = factor(default))
test_df  <- test_df %>% mutate(default = factor(default))
```

```{r}
metrics_class <- metric_set(accuracy)

metrics_prob <- metric_set(
  roc_auc,
  pr_auc,
  mn_log_loss,
  brier_class
)

eval_metrics <- function(df, model_name) {
  bind_rows(
    accuracy(df, truth = default, .class),
    roc_auc(df, truth = default, .pred, event_level = "second"),
    pr_auc(df,  truth = default, .pred, event_level = "second"),
    mn_log_loss(df, truth = default, .pred, event_level = "second"),
    brier_class(df, truth = default, .pred, event_level = "second")
  ) %>%
    mutate(model = model_name)
}
```

## Regularized logistic regression

```{r}
x_train <- model.matrix(default ~ . - 1, train_df)
y_train <- train_df$default

x_test <- model.matrix(default ~ . - 1, test_df)

cv_glmnet <- cv.glmnet(
  x_train,
  y_train,
  family = "binomial",
  type.measure = "auc"
)

glmnet_probs <- predict(cv_glmnet, x_test, s = "lambda.min", type = "response")[,1]

glmnet_preds <- test_df %>%
  mutate(
    .pred = glmnet_probs,
    .class = if_else(.pred >= 0.5, "1", "0") %>% factor(levels = levels(default))
  )
```

## Random Forest (ROC-AUC)

```{r}
rf_model <- ranger(
  default ~ .,
  data = train_df,
  probability = TRUE,
  num.trees = 500,
  mtry = floor(sqrt(ncol(train_df) - 1))
)

rf_probs <- predict(rf_model, test_df)$predictions[, "1"]

rf_preds <- test_df %>%
  mutate(
    .pred = rf_probs,
    .class = if_else(.pred >= 0.5, "1", "0") %>% factor(levels = levels(default))
  )
```

## Gradient Boosting (AUC)

```{r}
dtrain <- xgb.DMatrix(data = x_train, label = as.numeric(y_train) - 1)
dtest  <- xgb.DMatrix(data = x_test)

params <- list(
  objective = "binary:logistic",
  eval_metric = "auc",
  max_depth = 4,
  eta = 0.1,
  subsample = 0.8,
  colsample_bytree = 0.8
)

xgb_model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = 200,
  verbose = 0
)

xgb_probs <- predict(xgb_model, dtest)

xgb_preds <- test_df %>%
  mutate(
    .pred = xgb_probs,
    .class = if_else(.pred >= 0.5, "1", "0") %>% factor(levels = levels(default))
  )
```

## Results

```{r}
glmnet_results <- eval_metrics(glmnet_preds, "Logistic (regularized)")
rf_results     <- eval_metrics(rf_preds, "Random Forest")
xgb_results    <- eval_metrics(xgb_preds, "Gradient Boosting")

all_results <- bind_rows(
  baseline_results %>% mutate(model = "Baseline Logistic"),
  glmnet_results,
  rf_results,
  xgb_results
)

results_wide <- all_results %>%
  select(model, .metric, .estimate) %>%
  pivot_wider(names_from = .metric, values_from = .estimate)

results_wide
```

```{r}
all_results <- all_results %>%
  mutate(
    model = factor(
      model,
      levels = c(
        "Baseline Logistic",
        "Logistic (regularized)",
        "Random Forest",
        "Gradient Boosting"
      )
    )
  )

metric_disagreement_plot <- ggplot(all_results, aes(x = model, y = .estimate, color = .metric, group = .metric)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ .metric, scales = "free_y") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    title = "Metric Behavior as Model Complexity Increases",
    x = "Model (increasing complexity)",
    y = "Metric Value"
  )

saveRDS(metric_disagreement_plot, "../figures/metric_disagreement_plot")
saveRDS(glmnet_preds,             "../data/glmnet_preds.rds")
saveRDS(rf_preds,                 "../data/rf_preds.rds")
saveRDS(xgb_preds,                "../data/xgb_preds.rds")
```

More complex models improve ranking metrics (ROC-AUC, PR-AUC), but probability calibration did not consistently improve. In fact, the model with the strongest ranking performance, gradient boosting, produced worse calibrated probabilities, as reflected by a higher Brier score. This demonstrates that optimizing for discrimination can conflict with probability quality.
