---
title: "Metric Tradeoffs: Ranking vs Calibration"
format: html
---

## Goal
Explain why ROC-AUC / PR-AUC can improve while Brier score and log loss do not.

```{r}
baseline_preds <- readRDS("../data/baseline_preds.rds")
glmnet_preds <- readRDS("../data/glmnet_preds.rds")
rf_preds <- readRDS("../data/rf_preds.rds")
xgb_preds <- readRDS("../data/xgb_preds.rds")
metric_disagreement_plot <- readRDS("../figures/metric_disagreement_plot")

baseline_preds <- baseline_preds %>% mutate(default = factor(default, levels = c("0","1")))
glmnet_preds <- glmnet_preds %>% mutate(default = factor(default, levels = c("0","1")))
rf_preds <- rf_preds %>% mutate(default = factor(default, levels = c("0","1")))
xgb_preds <- xgb_preds %>% mutate(default = factor(default, levels = c("0","1")))
```

```{r}
preds_long <- bind_rows(
  baseline_preds %>% mutate(model = "Baseline Logistic"),
  glmnet_preds   %>% mutate(model = "Logistic (regularized)"),
  rf_preds       %>% mutate(model = "Random Forest"),
  xgb_preds      %>% mutate(model = "Gradient Boosting")
) %>%
  mutate(default = factor(default, levels = c("0","1")))

calibration_by_class <- function(df, n_bins = 10) {
  df_long <- df %>%
    mutate(
      y1 = as.numeric(default == "1"),
      y0 = as.numeric(default == "0"),
      p1 = .pred,
      p0 = 1 - .pred
    ) %>%
    select(model, y1, y0, p1, p0) %>%
    pivot_longer(
      cols = c(y1, y0, p1, p0),
      names_to = c(".value", "class"),
      names_pattern = "([yp])(\\d)"
    ) %>%
    mutate(class = if_else(class == "1", "Default = 1", "Default = 0"))

  df_long %>%
    mutate(bin = ntile(p, n_bins)) %>%
    group_by(model, class, bin) %>%
    summarise(
      n = n(),
      pred_mean = mean(p),
      obs_rate  = mean(y),
      .groups = "drop"
    )
}

cal_class_df <- calibration_by_class(preds_long, n_bins = 10)

ggplot(cal_class_df, aes(x = pred_mean, y = obs_rate, color = model)) +
  geom_abline(linetype = 2) +
  geom_line() +
  geom_point(aes(size = n), alpha = 0.9) +
  facet_wrap(~ class) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  theme_minimal() +
  labs(
    title = "Calibration Curves Faceted by Class",
    x = "Mean predicted probability",
    y = "Observed frequency",
    size = "Bin count"
  )

library(tidyverse)

# preds_long must have: model, default (factor "0","1"), .pred (P(default=1))

cal_long <- preds_long %>%
  mutate(
    y1 = as.numeric(default == "1"),
    y0 = as.numeric(default == "0"),
    p1 = .pred,
    p0 = 1 - .pred
  ) %>%
  select(model, y1, y0, p1, p0) %>%
  pivot_longer(
    cols = c(y1, y0, p1, p0),
    names_to = c(".value", "class"),
    names_pattern = "([yp])(\\d)"
  ) %>%
  mutate(
    class = if_else(class == "1", "Default = 1", "Default = 0")
  )

ggplot(cal_long, aes(x = p, y = y, color = model)) +
  geom_abline(linetype = 2) +
  geom_point(alpha = 0.03, size = 0.6) +
  geom_smooth(method = "loess", se = FALSE, span = 0.8) +
  facet_wrap(~ class) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  theme_minimal() +
  labs(
    title = "Calibration Curves (No Bins): Raw Outcomes + LOESS Smooth",
    x = "Predicted probability",
    y = "Actual outcome (0/1)"
  )

```

```{r}
metric_disagreement_plot
```

## Foundational Knowledge

1. Ranking metrics: ROC-AUC, PR-AUC

- Is a riskier person ranked above a safer one?

- They do not care about the actual probability values.

- They are insensitive to calibration.

Complex models are good at this because they carve complex decision boundaries and push predictions towards 0 or 1. As complexity increases, ranking metrics tend to improve.

2. Probability quality metrics: Log loss, Brier score

- These care about numerical honesty. If the model says 0.8, the event should happen ~80% of the time Overconfidence and underconfidence are punished

Complex models aren't as good at this because they can become overconfident and push predictions toward 0 or 1, improving separation but hurting calibration. This must be explicitly corrected for in complex models.

3. Threshold metrics: Accuracy, F1

- Imagine you have a dataset where 90% are class 0 and 10% are class 1. Suppose you build the dumbest model in the world that always predicts 0. Your accuracy would be 90%!

- Now imagine you have another model that correctly finds half of the positives but introduces false positives 10% of the time. Your accuracy would be (5 + 81)/100, or 86%. A 4% decrease despite this second model being far more useful for prediction.

## Findings

Accuracy is almost perfectly flat across all models. The dataset is imbalanced, so this matches expectations perfectly.

ROC-AUC slightly improves, with Gradient Boosting best. This tells you that the problem is already almost linearly separable in the feature space; there isn’t a ton of hidden nonlinear structure for complex models to exploit. Boosting is a ranking machine and logistic regression already captures most linear signal.

PR-AUC improves modestly. PR is harder to move and improvements reflect better minority-class focus.

Log loss is also flat, indicating probability quality did not meaningfully improve with complexity, especially since we didn't explicitly correct for this. This also matches expectations perfectly.

For Brier score, Random Forest produced the best calibrated probabilities, while Gradient Boosting produced the worst calibrated probabilities, even though Gradient Boosting had the best ranking metrics. This shows a clear tradeoff: best ranking vs worst calibration.

In conclusion, discrimination ≠ calibration ≠ decision quality, and class imbalance amplifies the difference between them.

If I were to recreate this experiment with balanced data, accuracy would actually become sensitive to model quality. I would probably see bigger accuracy difference between models. Accuracy would also track ROC-AUC more closely.

For PR-AUC, differences between models would shrink.

For Brier score, your model would need to be more calibrated to achieve a good Brier score. Calibration differences would become much more visible.

Metric disagreement would probably shrink a bit.

That's all for now! This was super informational and I had a lot of fun :)
